---
title: "Data Visualization Group Project"
author: Our Group
date: "February 11, 2016"
output: html_document
---

This is our first homework.

Data cleaning:


```{r echo=FALSE, results='hide',message=FALSE}


library(gdata)
library(ggplot2)
library(plyr)
library(MASS)
library(gcookbook)
library(rgl) 
library(Rmisc)
library(lattice)


## Set your own working directory!!!
setwd("C:/Users/Eartha/Desktop/DV Project 1")
#survey <- read.xls("Survey+Response.xlsx", stringsAsFactors = F)
#for windows user 
survey <- read.xls("Survey+Response.xlsx", stringsAsFactors = F,perl="C:\\Perl64\\bin\\perl.exe")


########################
#### DATA CLEANING ####
#######################

## Add ID variable.
survey <- cbind(1:nrow(survey), survey)

## Drop useless columns and rename the good ones.
survey <- survey[,colSums(is.na(survey))<nrow(survey)]
names(survey)
newNames <- c("id", "Waitlist", "Program", "Tools", "R - Data Manipulation", "Gender", "Prof. Editor", "R - Graphic Skills",
              "R - Adv. Multivariate Analysis Skills", "R - Reproducibility Skills", "Matlab Skills", "Github Skills")
names(survey) <- newNames

## Get list of all available "tools".
toolsList <- as.character(survey$Tools)
toolsList <- strsplit(toolsList, ",")
toolsList <- unique(trim(unlist(toolsList)))
toolsList
lg <- length(toolsList)

## Make binary variables for each tool.
test <- data.frame(survey$id, survey$Tools, stringsAsFactors = F)
names(test) <- c("id", "tools")
out <- data.frame(matrix(ncol=lg, nrow=0))
names(out) <- toolsList

make_row = function(x) {
  m <- c(rep(NA,lg))
  names(m) <- toolsList
  v = as.character(unique(trim(unlist(strsplit(x[1,2], ",")))))
  m[v] = "Yes"
  m
}

outFile <- ddply(.data=test, .variables=.(id), .fun=make_row)

surveyNew <- merge(survey, outFile, by="id")
surveyNew <- surveyNew[!(names(surveyNew) %in% c("Tools"))]

## Clean program options.
sort(unique(surveyNew$Program))

surveyNew[surveyNew$Program == "Ms in ds",]$Program <- "IDSE (master)"
surveyNew[surveyNew$Program == "MSDS",]$Program <- "IDSE (master)"
surveyNew[surveyNew$Program == "Data Science",]$Program <- "IDSE (master)"
surveyNew[surveyNew$Program == "QMSS (master)",]$Program <- "QMSS"
surveyNew[surveyNew$Program == "Data Science Certification",]$Program <- "IDSE (certificate)"
surveyNew[surveyNew$Program == "PhD Biomedical Informatics",]$Program <- "Ph.D."

#ggplot(surveyNew, aes(Program, fill=Program)) + geom_bar() +
#  geom_text(aes(label = format(..count.., digits=2, drop0trailing=TRUE), y= ..count.. ), stat= "count", vjust = -.5) +
#  theme(legend.position = "bottom")

## Clean text editor.
sort(unique(surveyNew$`Prof. Editor`))
surveyNew[surveyNew$`Prof. Editor` == "Any (20 years C++/Java experience)",]$`Prof. Editor` <- "Any"
surveyNew[grepl("sublime", surveyNew$`Prof. Editor`, ignore.case=T),]$`Prof. Editor` <- "Sublime"
surveyNew[surveyNew$`Prof. Editor` == "I used jupyter last semester",]$`Prof. Editor` <- "Jupiter"
surveyNew[surveyNew$`Prof. Editor` == "textwrangler",]$`Prof. Editor` <- "textWrangler"
surveyNew[surveyNew$`Prof. Editor` == "Text Wrangler",]$`Prof. Editor` <- "textWrangler"
surveyNew[surveyNew$`Prof. Editor` == "python",]$`Prof. Editor` <- "iPython"
surveyNew[surveyNew$`Prof. Editor` == "ipynb",]$`Prof. Editor` <- "iPython"
surveyNew[surveyNew$`Prof. Editor` == "Webstorm, pycharm",]$`Prof. Editor` <- "iPython"
surveyNew[surveyNew$`Prof. Editor` == "notepad++",]$`Prof. Editor` <- "Notepad++"
surveyNew[surveyNew$`Prof. Editor` == "xcode",]$`Prof. Editor` <- "XCode"
surveyNew[surveyNew$`Prof. Editor` == "vi/vim",]$`Prof. Editor` <- "Vi/Vim"
surveyNew[surveyNew$`Prof. Editor` == "haven't used any",]$`Prof. Editor` <- "None"


## Clean gender.
sort(unique(surveyNew$Gender))
surveyNew[surveyNew$Gender == "",]$Gender <- "he/him"
surveyNew[surveyNew$Gender == "doesn't matter",]$Gender <- "she/her"
surveyNew[surveyNew$Gender == "he/him",]$Gender <- "Male"
surveyNew[surveyNew$Gender == "she/her",]$Gender <- "Female"
```



#Summary Statistics:




```{r, echo=FALSE, fig.width=10, fig.height=8}
#bar chart for Prof Editor
ggplot(surveyNew, aes(`Prof. Editor`, fill=`Prof. Editor`)) + geom_bar() +
  geom_text(aes(label = format(..count.., digits=2, drop0trailing=TRUE), y= ..count.. ), stat= "count", vjust = -.5) +
  theme(legend.position = "bottom")
```



```{r, echo=FALSE}

#waitlist statstic
waitlist=rbind(table(surveyNew$Waitlist))

```



```{r, echo=FALSE}
#pie chart for waitlist
waitlist <- table(surveyNew$Waitlist)
lbls <- paste(names(waitlist), "\n", waitlist, sep="")
pie(waitlist, col= c("blue","red"),labels = lbls, main="Waitlist")

```



```{r, echo=FALSE, fig.width=10, fig.height=8}
#Program statstic
program <- rbind(table(surveyNew$Program))
bp_program=barplot(as.matrix(program),main="Students by Program",ylab="numbers",cex.names=0.7,cex.main=1.4,beside = TRUE, col=rainbow(7))
text(x=bp_program, y=program, labels=round(program,0), pos=3, xpd=NA)
```

Students from Data Science program are the majority since this is the required course and follow by Statistics program. Interestingly we have 1 student from Applied Math and 3 students from Ph.D. 


```{r, echo=FALSE}
#Gender Statistic
gender <-rbind(table(surveyNew$Gender))
```



```{r, echo=FALSE}
#pie chart for Gender
gender <- table(surveyNew$Gender)
lbls <- paste(names(gender), "\n", gender, sep="")
pie(gender, col= c("red","blue"),labels = lbls, main="Gender")

```


Male students take up 71% of the class 



``````{r echo=FALSE, results='hide',message=FALSE}

#create matrix for skills with proficiency
skills <- cbind (table(surveyNew$`R - Data Manipulation`),
                 table(surveyNew$`R - Graphic Skills`),
                 table(surveyNew$`R - Adv. Multivariate Analysis Skills`),
                 table(surveyNew$`R - Reproducibility Skills`),
                 table(surveyNew$`Matlab Skills`), 
                 table(surveyNew$`Github Skills`))
#Skills Statistic with proficiency
colnames(skills) <- c("R - Data Manipulation","R - Graphic","R - Adv.Multivariate","R - Reproducibility","Matlab Skills","Github Skills")
skills

```



```{r, echo=FALSE,fig.width=10,fig.height=8}

#bar chart for skills with proficiency
colours <- c("red", "orange", "blue", "yellow")
bp_skills=barplot(as.matrix(skills), main="skills statistic", ylab = "Numbers",cex.names=0.8, cex.lab = 0.8, cex.main = 1.4, beside=TRUE, col=colours)
legend("topright", c("A little","Confident","Expert","None"), cex=0.8, bty="n", fill=colours)
text(x=bp_skills, y=skills, labels=round(skills,0), pos=3, xpd=NA)
```  


Half students are confident in R-data manipulation. However, in terms of R-Graphic,R-Adv.Multivariate,R-Reproductibility, Matlab, and github, most students have little or no experience. 



```{r, echo=FALSE, fig.width=10,fig.height=8}
#change "yes" response to 1 and none response to 0
tool_exp <- surveyNew[,12:31]

tool_exp[tool_exp[,]== "Yes"] <- "1"
tool_exp[is.na(tool_exp[,])] <- 0
#transform charactor structure to numeric
tool_exp=as.matrix(sapply(tool_exp, as.numeric))
```



``````{r echo=FALSE, fig.width=10, fig.height=8, results='hide',message=FALSE}
#statistic table for tools
#colSums() calcuates total students that have experience for each tool
tool_exp <- rbind(colSums(tool_exp)) 
tool_exp
tool_exp_name=names(surveyNew[,12:31])

#adjust the margins
par(mar=c(5.1, 13 ,4.1 ,2.1))
bp_tool_exp=barplot(as.matrix(tool_exp),main="Experience with Tools Statistic",
                    horiz=TRUE,legend=TRUE,xlab="numbers",names.arg=tool_exp_name, 
                    las=1,cex.names=0.8,xlim=c(0,100),cex.main=1.4,beside = TRUE, col=rainbow(20))

 
#adjust font size and numbers positon
text(x=tool_exp, y=bp_tool_exp,cex=0.8,labels=round(tool_exp,0), pos=4, xpd=NA)
#adjust the margins
par(mar=c(5.1, 13 ,4.1 ,2.1))
```


As majority students are from Data Science and Statistics program, students tend to have experience in R, RStudio, Python and Excel as these are the most popular tools for data analysis and statistics. 




# Conditional Distribution

```{r echo=FALSE, results='hide',message=FALSE}
####################################
##                                ##
##    Conditional Distributions   ##
##                                ##
####################################

# Create a new survey table that convert 
#convert "yes" in tools to 1 and calculated the sum of "yes" as the tool experience score.
#create the tool_coverted table.

tool_numeric<- surveyNew[,c(12:31)]
tool_numeric[tool_numeric[,]== "Yes"] <- "1"
tool_numeric[is.na(tool_numeric[,])] <- 0
tool_numeric=as.matrix(sapply(tool_numeric, as.numeric))

tool_converted <- cbind(tool_numeric,tool_experience_level = rowSums(tool_numeric))
tool_converted

#convert the "None", "A little", "Confident","Expert" to number 1, 2, 3, 4 respectively.
#create the experience_converted table


programming_experience<- surveyNew[,c("R - Data Manipulation","R - Graphic Skills","R - Adv. Multivariate Analysis Skills","R - Reproducibility Skills","Matlab Skills","Github Skills")]


programming_experience$`R - Data Manipulation`[programming_experience$`R - Data Manipulation` == "None"] <- 0
programming_experience$`R - Data Manipulation`[programming_experience$`R - Data Manipulation` == "A little"] <- 1
programming_experience$`R - Data Manipulation`[programming_experience$`R - Data Manipulation` == "Confident"] <- 2
programming_experience$`R - Data Manipulation`[programming_experience$`R - Data Manipulation` == "Expert"] <- 3

programming_experience$`R - Graphic Skills`[programming_experience$`R - Graphic Skills` == "None"] <- 0
programming_experience$`R - Graphic Skills`[programming_experience$`R - Graphic Skills` == "A little"] <- 1
programming_experience$`R - Graphic Skills`[programming_experience$`R - Graphic Skills` == "Confident"] <- 2
programming_experience$`R - Graphic Skills`[programming_experience$`R - Graphic Skills` == "Expert"] <- 3

programming_experience$`R - Adv. Multivariate Analysis Skills`[programming_experience$`R - Adv. Multivariate Analysis Skills` == "None"] <- 0
programming_experience$`R - Adv. Multivariate Analysis Skills`[programming_experience$`R - Adv. Multivariate Analysis Skills` == "A little"] <- 1
programming_experience$`R - Adv. Multivariate Analysis Skills`[programming_experience$`R - Adv. Multivariate Analysis Skills` == "Confident"] <- 2
programming_experience$`R - Adv. Multivariate Analysis Skills`[programming_experience$`R - Adv. Multivariate Analysis Skills` == "Expert"] <- 3

programming_experience$`R - Reproducibility Skills`[programming_experience$`R - Reproducibility Skills` == "None"] <- 0
programming_experience$`R - Reproducibility Skills`[programming_experience$`R - Reproducibility Skills` == "A little"] <- 1
programming_experience$`R - Reproducibility Skills`[programming_experience$`R - Reproducibility Skills` == "Confident"] <- 2
programming_experience$`R - Reproducibility Skills`[programming_experience$`R - Reproducibility Skills` == "Expert"] <- 3


programming_experience$`Matlab Skills`[programming_experience$`Matlab Skills` == "None"] <- 0
programming_experience$`Matlab Skills`[programming_experience$`Matlab Skills` == "A little"] <- 1
programming_experience$`Matlab Skills`[programming_experience$`Matlab Skills` == "Confident"] <- 2
programming_experience$`Matlab Skills`[programming_experience$`Matlab Skills` == "Expert"] <- 3

programming_experience$`Github Skills`[programming_experience$`Github Skills` == "None"] <- 0
programming_experience$`Github Skills`[programming_experience$`Github Skills` == "A little"] <- 1
programming_experience$`Github Skills`[programming_experience$`Github Skills` == "Confident"] <- 2
programming_experience$`Github Skills`[programming_experience$`Github Skills` == "Expert"] <- 3

programming_experience=as.matrix(sapply(programming_experience, as.numeric))
programming_experience

experience_converted <- cbind(programming_experience,programming_experience_level = rowSums(programming_experience))
experience_converted


#Create a new survey table "surveyNew_numeric" that convert strings to numeric
rest_of_surveyNew <- surveyNew[,c("id", "Waitlist", "Program", "Gender", "Prof. Editor")]
surveyNew_numeric <- cbind(rest_of_surveyNew, experience_converted, tool_converted)

surveyNew_numeric





```



```{r, echo=FALSE，fig.width=8, fig.height=8}
# Bar Chart 2
barchart2 <- ggplot(surveyNew_numeric,aes(x=Program,y=programming_experience_level,fill=Gender))+
  geom_bar(stat="identity",position="dodge")+
  geom_text(aes(label=tool_experience_level), vjust=1.5, colour="white",
            position=position_dodge(.9),size=3)

barchart2+  xlab("Programs") + ylab("Programming Confidence Level") + ggtitle("Compare programming confidence level\n among programs group by gender ")
```


Programming and Analytical Experiece Answer and corresponding scores:
None = 0     A little = 1     Confident = 2      Expert = 3

This Bar Graph shows the programming confidence level of males and females grouped 
by programs. IDSE masters' males students has the highest programming confidence level
and Applied Math's male students has the lowest confidence level. In general, male 
students programming confidence level is higher than females'.







```{r echo=FALSE, results='hide',message=FALSE}
#make a copy of the data
surveyNew_numeric1 <- surveyNew_numeric
```



```{r, echo=FALSE, fig.width=8, fig.height=7}
#library(Rmisc)
#Bar Chart 3
# R vs tool_experience_level
surveyNew_numeric1$R <- factor(surveyNew_numeric1$R)
levels(surveyNew_numeric1$R)
surveyNew_numeric1$R <- revalue(surveyNew_numeric1$R, c("1"="Experience with R","0"="No experience with R"))
R_vs_experience <- ggplot(surveyNew_numeric1, aes(x=tool_experience_level, fill=R)) +    geom_histogram(position="identity", alpha=0.4)
R_vs_experience1 <- R_vs_experience + ggtitle("Compare the Tool skill Count bewteen students \n who had experience with R and who didn't ") + xlab("Tool Skill Count")


# Density Curve 1
#library(MASS) #For the data set
#Density Curves from grouped data
#Density curve  R
surveyNew_numeric2 <- surveyNew_numeric
surveyNew_numeric2$R <- factor(surveyNew_numeric2$R)
surveyNew_numeric2$R <- revalue(surveyNew_numeric1$R, c("1"="Experience with R","0"="No experience with R"))
#Map R to colour
dentisy_curve1<-ggplot(surveyNew_numeric2, aes(x=tool_experience_level, colour=R)) + geom_density()
#Map R to fill and make the fill semitransparent by setting alpha
density_curve1<-ggplot(surveyNew_numeric2, aes(x=tool_experience_level, fill=R)) + geom_density(alpha=.3)
density_curve_combine<-density_curve1 + ggtitle("Density Curve") + xlab("Tool Skill Count")

multiplot(R_vs_experience1,density_curve_combine)

```


The two graphs above shows the bar chart and density curve of tool skill count.
The meanning of tool skill count: if a student have one tool still, then the tool skill count is 1, if the student have two tool skills then the tool skill count is 2, etc.From the bar chart we can see, there are around 16 students who had experience with R that have a tool skill count of 6 and there are around 7 students who don't have experience with R have the tool skill count of 4. From the density curve it is easily to see that the peak of the curve which shows students who don't have experience with R is 0.25. It means that 25% of the students who don't have experience with R has four tool skills. 13% of the students who had experience with R have 6 tool skills.





```{r, echo=FALSE}
# Frequency Polygon
# tool experience level
fre_poly<-ggplot(surveyNew_numeric, aes(x=tool_experience_level)) + geom_freqpoly(binwidth=4)
# programming experience level
fre_poly<-ggplot(surveyNew_numeric, aes(x=programming_experience_level)) + geom_freqpoly(binwidth=4)
fre_poly + ggtitle("Frequeny Poygon: distribution of all the \n students' programming confidence level") + xlab("Programming Confidence Level")

```


The above frequency poygon graph shows the counts of students at each programming confidence levels. The programming confidence level for "R, data manipulation and modeling", "R, graphic basics (base, lattice, grid etc. )", " R, advanced (multivariate data analysis, e.g. spatiotemporal data, visualization and modeling) ","Reproducible documentation with R (e.g. R Markdown) ", "Matlab, data manipulation, analysis, visualization and modeling" and "Github" is "None = 0 A little = 1 Confident = 2 Expert = 3". Each student's programming confidence level is the sum of confidence level of each programming and analytical experience. The peak of the curve is the point of(x=7, y=56), which means there are 56 students who did the survey that have the programming confidence level of 56. 





```{r, echo=FALSE, fig.width=8,fig.height=7}
#box plot
#adding means to box plot
# tool experience level vs program
library(MASS) # For the data set
#The base plot
boxplot<-ggplot(surveyNew_numeric, aes(x=factor(Program), y=tool_experience_level,
  fill=Program)) + geom_boxplot() + stat_summary(fun.y="mean", 
  geom="point", shape=23, size=3, fill="white")
boxplot + ggtitle("Box Plot of tool skill count grouped by programs") +
  xlab("Programs") + ylab("Tool Skill Count")

```


The above graph shows the Box Plots of tool skill count of each program. It shows clearly the distribution of student's tool skill count for each program. 



```{r, echo=FALSE}
#Violin Plot 1
#library(gcookbook) # For the data set
# Base plot 
# gender vs tool_experience_level
p <- ggplot(surveyNew_numeric, aes(x=Gender, y=tool_experience_level))
#p + geom_violin()
#Violin with median point
violin_plot <- p + geom_violin() + geom_boxplot(width=.1, fill="Black", outlier.colour=NA) +   
  stat_summary(fun.y=median, geom="point", fill="white", shape=21, size=2.5) 
violin_plot + ggtitle("Violin Plot of Tool Skill Count grouped by Gender") + ylab("Tool Skill Count")

```


The above graph compares the violion plot of tool skill count between male and female studnts. The little white point inside each violin plot is the median. The difference of the median for male and female is not obvious. 


```{r, echo=FALSE}
#Violin Plot 2
# gender vs programming experience level
p <- ggplot(surveyNew_numeric, aes(x=Gender, y=programming_experience_level))
#p + geom_violin()
#Violin with median point
violin_plot2 <- p + geom_violin() + geom_boxplot(width=.1, fill="black", outlier.colour=NA) +   
  stat_summary(fun.y=median, geom="point", fill="white", shape=21, size=2.5) 
violin_plot2 + ggtitle("violin Plot of Programming Confidence Level grouped by gender") +
  ylab("Programming Confidence Level")


```


The above graph compares the violin plot of student's programming confidence level between male and female students. The same with the previous graph,the little white point inside each violin plot is the median of programming confidence level. Even though the shape of the violin plot are different, the median is almsot the same. 



```{r, echo=FALSE}
# Density Plot of Two-doimensional Data

#Density Plot of Two Dimensional Data 1
# The base plot
m <- ggplot(surveyNew_numeric, aes(x=programming_experience_level, y=tool_experience_level))
density1<-m  + geom_point() + stat_density2d()
density1 + ggtitle("Density Plot of Two dimensional Data") + xlab("Programming Confidence Level") + ylab("Tool Skill Count")

```


The density plot of two dimensional data graph above shows the relationship between tool skill count and programming confidence level. From the graph we can see that the more tool skills a student has, the more confident the student is at programming. The majority of students' programming confidence level is less than 10 and the tool skill count is less than 12. 


